# **Phân tích ANOVA 2 yếu tố kiểu RCBD trong R** 

### **Bước 1: Import dữ liệu**

```{r}
library(readxl)
data_anova <- read_excel("test.xlsx", 
                         sheet = "anova-twoway", range = "D5:F53")
data_anova <- as.data.frame(data_anova)
data_anova$diet <- as.factor(data_anova$diet)
data_anova$time <- as.factor(data_anova$time)
data_anova -> my_data

my_data
```

```{r}
# Bố trí thí nghiệm
table(data_anova$diet, data_anova$time)
```

#### <span style="color:red">Bước 2: Giả thuyết cho ANOVA 2 yếu tố</span>

Nguồn: <https://statistics.laerd.com/spss-tutorials/two-way-anova-using-spss-statistics.php>

* Assumption #1: Your dependent variable should be measured at the continuous level $\Rightarrow$ <mark style="background-color: #FFFF00">OK</mark> 

* Assumption #2: Your two independent variables should each consist of two or more categorical, independent groups $\Rightarrow$ <mark style="background-color: #FFFF00">OK</mark> 

* Assumption #3: You should have independence of observations, which means that there is no relationship between the observations in each group or between the groups themselves. $\Rightarrow$ <mark style="background-color: #FFFF00">OK</mark> 

* Assumption #4: There should be no significant outliers. 

* Assumption #5: Your dependent variable should be approximately normally distributed for each combination of the groups of the two independent variables. 

* Assumption #6: There needs to be homogeneity of variances for each combination of the groups of the two independent variables. Again, whilst this sounds a little tricky, you can easily test this assumption in SPSS Statistics using Levene’s test for homogeneity of variances.


### **Bước 2: Đánh giá mức độ phân bố chuẩn**

Check the normality assumpttion

<span style="color: blue">**Cách 1**</span>

```{r, message=FALSE, warning=FALSE}
library(car)
leveneTest(stability ~ diet*time, data = my_data)
```

From the output above we can see that the p-value is (0.6936) not less than the significance level of 0.05. This means that there is no evidence to suggest that the variance across groups is statistically significantly different. Therefore, we can assume the homogeneity of variances in the different treatment groups.

**Tạm dịch: Kết quả cho thấy p-value là 0.6936 lớn hơn 0.05 (giả thuyết cho là có sự phân bố không chuẩn - heterogeneity). Do đó bộ dataset này có sự phân bố chuẩn (homogeneity) trong sự khác biệt giữa các nghiệm thức.**

<span style="color: blue">**Cách 2**</span>

**Cần tính anova trước để có data vẽ đồ thị qqplot**

```{r, message=FALSE, warning=FALSE}
# Compute two-way ANOVA test
res.aov2 <- aov(stability ~ diet + time, data = my_data)
# summary(res.aov2)
# anova(res.aov2)

# Compute two-way ANOVA test with interaction effect
res.aov3 <- aov(stability ~ diet + time + diet:time, data = my_data)
# anova(res.aov3)
```

**Vẽ đồ thị**

```{r, message=FALSE, warning=FALSE}
plot(res.aov3, 1) ## Homogeneity of variances
plot(res.aov3, 2) ## Check the normality assumpttion
```

Phát hiện các data point 6, 13, 15 là outlier, *có thể* loại ra để làm dataset phân bố chuẩn hơn.

<span style="color: blue">**Cách 3**</span>

```{r, message=FALSE, warning=FALSE}
# Extract the residuals
aov_residuals <- residuals(object = res.aov3)
# Run Shapiro-Wilk test
shapiro.test(x = aov_residuals) 
```

p-value từ test Shapiro-Wilk normality cho thấy nhỏ hơn 0.05 (giả thuyết là phân bố chuẩn), do đó về mặt ý nghĩa thống kê thì bộ dataset này có phân bố chuẩn.

### **Bước 3: Khảo sát đặc điểm dữ liệu**

**Histogram theo `time`**

```{r, message=FALSE, warning=FALSE}
library(lattice)
histogram( ~ stability | time, data = my_data,
           xlab = "Stability (%)", type = "density",
           breaks = seq(from = 80, to = 95, by = 0.4),
           panel = function(x, ...) {
             panel.histogram(x, ...)
             panel.mathdensity(dmath = dnorm, col = "black",
                               args = list(mean=mean(x),sd=sd(x)))
           } )
```

**Box plot with multiple groups**

```{r, message=FALSE, warning=FALSE}
library(ggpubr)
ggboxplot(my_data, x = "diet", y = "stability", color = "time",
          palette = c("#00AFBB", "#E7B800"))
```

**Line plots with multiple groups**

```{r, message=FALSE, warning=FALSE}
# Add error bars: mean_se
# (other values include: mean_sd, mean_ci, median_iqr, ....)
library("ggpubr")
ggline(my_data, x = "diet", y = "stability", color = "time",
       add = c("mean_sd", "dotplot"),
       palette = c("#00AFBB", "#E7B800"))
```

**Box plot with two factor variables**

```{r, message=FALSE, warning=FALSE, fig.width=9}
oldpar <- par(no.readonly = TRUE)
par(mar = c(6, 7, 1, 6))
boxplot(stability ~ diet * time, data = my_data, frame = TRUE, 
        col = c("#00AFBB", "#E7B800"), horizontal = TRUE, las = 1,
        axisnames = TRUE, ylab = "", xlab = "Stability (%)")
par(oldpar)
```

<span style="color: red">**Nếu muốn vẽ `boxplot` theo thứ tự các cột thì cần `reorder` cột factor `diet` theo `stability`**</span>

```{r, message=FALSE, warning=FALSE, fig.width=9}
my_data$diet <- reorder(my_data$diet, my_data$stability, decreasing = TRUE) 
my_data$time <- reorder(my_data$time, my_data$stability, decreasing = TRUE) 

oldpar <- par(no.readonly = TRUE)
par(mar = c(6, 7, 1, 6))
boxplot(stability ~ diet * time, data = my_data, frame = TRUE, 
        col = c("#00AFBB", "#E7B800"), horizontal = TRUE, las = 1,
        axisnames = TRUE, ylab = "", xlab = "Stability (%)")
par(oldpar)
```

**Two-way interaction plot**

```{r, message=FALSE, warning=FALSE, fig.width=9}
interaction.plot(x.factor = my_data$diet, trace.factor = my_data$time, 
                 response = my_data$stability, fun = mean, 
                 type = "b", legend = TRUE, 
                 xlab = "Diet", ylab ="Stability (%)",
                 pch = c(1, 19), col = c("#00AFBB", "#E7B800"),
                 ylim = c(50, 100))
```

### **Bước 4: Phân tích ANOVA 2 yếu tố CRD**

**Tính p-value**

```{r, message=FALSE, warning=FALSE}
# Compute two-way ANOVA test
res.aov2 <- aov(stability ~ diet + time, data = my_data)
# summary(res.aov2)
anova(res.aov2)

# Compute two-way ANOVA test with interaction effect
res.aov3 <- aov(stability ~ diet + time + diet:time, data = my_data)
anova(res.aov3)
```

**Phân hạng**

```{r, message=FALSE, warning=FALSE}
library(agricolae)
LSD.test(res.aov2, c("diet", "time"), console = TRUE) 

duncan.test(res.aov2, c("diet", "time"), console = TRUE) 

HSD.test(res.aov2, c("diet", "time"), console = TRUE) 
```

#### Tài liệu tham khảo

1.  `http://www.sthda.com/english/wiki/two-way-anova-test-in-r`

2.  `https://stackoverflow.com/questions/43123462/how-to-obtain-rmse-out-of-lm-result`

3.  `https://online.stat.psu.edu/stat501/lesson/2/2.6`

4.  `https://rcompanion.org/handbook/G_14.html`

5.  RMSE (Root Mean Square Error) `https://agronomy4future.org/?p=15930`

6.  `https://stats.stackexchange.com/questions/445200/coefficient-of-variation-for-beween-groups`

#### Một số kỹ thuật xử lý ANOVA 2 yếu tố


